---
title: "Analysis of natural disasters: effects on health and the economy"
author: "Francisco Javier Carela Ferrer"
date: "13/10/2020"
output: html_document
---
# Synopsis
This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.
To examine which weather events are the most most harmful with respect to population health, this analysis looks at the rates of injury, death, and property damage for all events since 1980, limited to those for which there are at least 500 occurences since 1980.

# Variables description

Using [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) we can check that:

 - *FATALITIES*: Number of deaths.
 - *INJURIES*: Number of injuries.
 - *PROPDMG*: Property damags.
 - *PROPDMGEXP*:  Units for property damage value.
 - *CROPDMG*: Crop damages.
 - *CROPDMGEXP*: Units for crop damage value.
 
 Events variable:
 
  - *EVTYPE*: Eeather event (Tornados, Wind, Snow, Flood, etc..).
  
# Data Processing

First of all , we are going to use the download.file() function to extract our dataset:
```{r}
#download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2","storms.csv.bz2")
```

The bz2 format is a zip format, but we can read it using read.cvs(), it doesn't works using unz().

```{r}
Storm_data<-read.csv('repdata_data_StormData.csv.bz2',stringsAsFactors=F)
```

# Data exploring

```{r echo = TRUE}
str(Storm_data)
```


As we are interested in variables related to health and the economy in relation to storms, we are going to carry out transformations in our data set:
```{r}
Storm_data$BGN_DATE <- as.Date(Storm_data$BGN_DATE,format="%m/%d/%Y")
```

And now we can subsetting from 1980:
```{r echo = TRUE}
library(lubridate)
Storm_data_sub <- Storm_data[year(Storm_data$BGN_DATE)>=1980,]
vars <- c( "EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")
storm <- Storm_data_sub[, vars]
dim(storm)
```
We can check the variables *PROPDMGEXP* and *CROPDMGEXP*, for Checkingthe values for variables that represent units od dollars.

```{r}
sort(table(storm$PROPDMGEXP), decreasing = TRUE)[1:10]

```
```{r}
sort(table(storm$CROPDMGEXP), decreasing = TRUE)[1:10]
```
 And we can check if we have any NA value:
```{r echo = TRUE}
library(dplyr)
summarise_all(storm, funs(sum(is.na(.))))
```

# Data transformation

First of all, we can put together some values in *EVTYPE*: 
```{r}
storm$EVTYPE[grep("BLIZZARD|WINTER",storm$EVTYPE)] <- "BLIZZARD"
storm$EVTYPE[grep("FLOOD|FLDG|FLD|URBAN|FLOOO",storm$EVTYPE)]<-"FLOOD"
storm$EVTYPE[grep("FREEZING RAIN|RAIN",storm$EVTYPE)] <- "RAIN"
storm$EVTYPE[grep("ICE|ICY",storm$EVTYPE)]<- "ICE"
storm$EVTYPE[grep("HURRICANE|TYPHOON",storm$EVTYPE)] <- "HURRICANE"
storm$EVTYPE[grep("THUNDER|TSTM",storm$EVTYPE)] <- "THUNDER STORM"
storm$EVTYPE[grep("LIGHTN|LIGHTI|LIGNT",storm$EVTYPE)] <- "LIGHTNING"
storm$EVTYPE[grep("WIND|WND",storm$EVTYPE)] <- "WIND"
storm$EVTYPE[grep("TORN|FUNNEL",storm$EVTYPE)] <- "TORNADO"
storm$EVTYPE[grep("HEAT|HOT|HIGH TEMP",storm$EVTYPE)] <- "HEATWAVE"
storm$EVTYPE[grep("SWELL|SEA|TIDE",storm$EVTYPE)] <- "SEAS"
storm$EVTYPE<-as.factor(storm$EVTYPE)
```

There is some mess in units, so we transform those variables in one unit (dollar) variable by the rule following here  [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) :


 - *K* or *k*: **thousand** dollars ($10^3$)
 - *M* or *m*: **million** dollars ($10^6$)
 - *B* or *b*: **billion** dollars ($10^9$)
 - the rest would be consider as dollars
 
```{r}
storm$PROPDMGEXP <- as.character(storm$PROPDMGEXP)
storm$PROPDMGEXP[is.na(storm$PROPDMGEXP)] <- 0 # NA's considered as dollars
storm$PROPDMGEXP[!grepl("K|M|B", storm$PROPDMGEXP, ignore.case = TRUE)] <- 0 # everything exept K,M,B
storm$PROPDMGEXP[grep("K", storm$PROPDMGEXP, ignore.case = TRUE)] <- "3"
storm$PROPDMGEXP[grep("M", storm$PROPDMGEXP, ignore.case = TRUE)] <- "6"
storm$PROPDMGEXP[grep("B", storm$PROPDMGEXP, ignore.case = TRUE)] <- "9"
storm$PROPDMGEXP <- as.numeric(as.character(storm$PROPDMGEXP))
storm$property.damage <- storm$PROPDMG * 10^storm$PROPDMGEXP

storm$CROPDMGEXP <- as.character(storm$CROPDMGEXP)
storm$CROPDMGEXP[is.na(storm$CROPDMGEXP)] <- 0 # NA's considered as dollars
storm$CROPDMGEXP[!grepl("K|M|B", storm$CROPDMGEXP, ignore.case = TRUE)] <- 0 # everything exept K,M,B is dollar
storm$CROPDMGEXP[grep("K", storm$CROPDMGEXP, ignore.case = TRUE)] <- "3"
storm$CROPDMGEXP[grep("M", storm$CROPDMGEXP, ignore.case = TRUE)] <- "6"
storm$CROPDMGEXP[grep("B", storm$CROPDMGEXP, ignore.case = TRUE)] <- "9"
storm$CROPDMGEXP <- as.numeric(as.character(storm$CROPDMGEXP))
storm$crop.damage <- storm$CROPDMG * 10^storm$CROPDMGEXP
```
 
We transform the values *PROPDMGEXP* and *PROPDMGEXP* because they have the value *n* in  $x^n$ , so we change the leters for the values value to $3$, $6$ and $9$ respectively. And know, we can see the results table:

```{r echo = TRUE}
sort(table(storm$property.damage), decreasing = TRUE)[1:5]

```

```{r echo = TRUE}
sort(table(storm$crop.damage), decreasing = TRUE)[1:5]

```

# Results 

Finally we can join fatalities and injuries in one table:

```{r echo = TRUE}
library(dplyr)
library(plyr)
# aggregate FATALITIES and INJURIES by type of EVTYPE
agg.fatalites.and.injuries <- ddply(storm, .(EVTYPE), summarize, Total = sum(FATALITIES + INJURIES,  na.rm = TRUE))
agg.fatalites.and.injuries$type <- "fatalities and injuries"
  
# aggregate FATALITIES by type of EVTYPE
agg.fatalities <- ddply(storm, .(EVTYPE), summarize, Total = sum(FATALITIES, na.rm = TRUE))
agg.fatalities$type <- "fatalities"

# aggregate INJURIES by type of EVTYPE
agg.injuries <- ddply(storm, .(EVTYPE), summarize, Total = sum(INJURIES, na.rm = TRUE))
agg.injuries$type <- "injuries"

# combine all
agg.health <- rbind(agg.fatalities, agg.injuries)
agg.health_top<-head(agg.health[order(agg.health$Total, decreasing = TRUE),],n =20L)

health.by.EVTYPE <- join (agg.fatalities, agg.injuries, by="EVTYPE", type="inner")
health_top<-head(health.by.EVTYPE[order(health.by.EVTYPE$Total, decreasing = TRUE),], n =20L)
health_top
```

And doing the same with the economic variables:
```{r echo = TRUE}
# aggregate PropDamage and CropDamage by type of EVTYPE
agg.propdmg.and.cropdmg <- ddply(storm, .(EVTYPE), summarize, Total = sum(property.damage + crop.damage,  na.rm = TRUE))
agg.propdmg.and.cropdmg$type <- "property and crop damage"

# aggregate PropDamage by type of EVTYPE
agg.prop <- ddply(storm, .(EVTYPE), summarize, Total = sum(property.damage, na.rm = TRUE))
agg.prop$type <- "property"

# aggregate INJURIES by type of EVTYPE
agg.crop <- ddply(storm, .(EVTYPE), summarize, Total = sum(crop.damage, na.rm = TRUE))
agg.crop$type <- "crop"

# combine all
agg.economic <- rbind(agg.prop, agg.crop)
agg.economic_top<-head(agg.economic[order(agg.economic$Total, decreasing = TRUE),],n =20L)


economic.by.EVTYPE <- join (agg.prop, agg.crop, by="EVTYPE", type="inner")
economic_top<-head(economic.by.EVTYPE[order(economic.by.EVTYPE$Total, decreasing = TRUE),],n =20L)
economic_top
```

## Answering the questions

### Across the United States, which types of events (as indicated in the \color{red}{\verb|EVTYPE|}EVTYPE variable) are most harmful with respect to population health?



```{r}
library(ggplot2)
library(ggthemes)
health.plot <- ggplot(agg.health_top, aes(x = EVTYPE, y = Total, fill = type)) + geom_bar(stat = "identity") + coord_flip() + xlab("Event Type") + ylab("Number of health impact") +
  ggtitle("Weather event types impact") + scale_fill_manual(values=c("red","black")) + theme_economist()
 
print(health.plot)  
```
 
***Tornadoes are the most damaging health events for our data set***

### Across the United States, which types of events have the greatest economic consequences?
 
```{r}
economic.plot <- ggplot(agg.economic_top, aes(x = EVTYPE, y = Total, fill = type)) + geom_bar(stat = "identity") + coord_flip() + xlab("Event Type") + ylab("Number of economic impact") +
  ggtitle("Weather event types impact") + scale_fill_manual(values=c("red","black")) + theme_economist()
 
print(economic.plot) 
```
 

***Thunder Storms  and Floods are the most economically damaging events for our data set***




*Francisco Javier Carela Ferrer*